---
title: "Customer personality analysis: Data visualization"
author: "Lim Jia Qi"
date: "5/24/2022"
output: 
  prettydoc::html_pretty:
    toc: TRUE
    theme: hpstr
    highlight: vignette
---

To render this Rmarkdown as nice theme html document, install `prettydoc` library and import it into R before knitting.

# Load library packages
```{r load_library, warning=FALSE}
library(tidyverse)  # Multiple packages for data analysis
library(visdat)     # visualize missing values
library(mice)       # multiple imputation of missing values
library(knitr)      # for tidy table display
library(kableExtra)  # for html table display
library(GGally)     # for scatter plot matrix
library(ggmosaic)   # mosaic plot
library(reshape2)   # data wrangling
library(ggforce)    # for parallet set plots
```

# Data acquisition 
Load the data into R workspace. The data can be downloaded from [Kaggle](https://www.kaggle.com/datasets/imakash3011/customer-personality-analysis).
```{r read_data, cache=TRUE}
# Set the working directory to where the data file is located
setwd("~/ml projects/customer segmentation")
cust_data = read.csv("marketing_campaign.csv", sep = "\t")
```

There are `r ncol(cust_data)` columns (variables) and `r nrow(cust_data)` rows (observations). Each observation corresponds to a customer.
```{r data_structure, echo = FALSE}
print(colnames(cust_data))
cust_data %>% str()
summary(cust_data)
```
From the output of `summary()` function, it is noted that there are missing values for the `Income` attribute. Aside from that, there are some other things to keep in mind:
1. Feature like `ID` is probably not useful as it does not encode any information about a customer. I will keep it in the `DataFrame` variable but not using it for any analysis.
2. Feature like `Z_CostContact` and `Z_Revenue` are not useful attributes as they have zero variance.
```{r variance, echo=FALSE}
cat("variance of ", colnames(cust_data)[ncol(cust_data)-2], " is: ", var(cust_data$Z_CostContact))
cat("variance of ", colnames(cust_data)[ncol(cust_data)-1], " is: ", var(cust_data$Z_Revenue))
```

3. Some feature engineering is required for some other features like `Dt_Customer` and `Year_Birth`. 
4. Feature restructuring from `character` to `factor` for features such as `Education` and `Marital Status`.

# Feature engineering

## Removal of zero variance variables

```{r set_date}
cust_data_copy = cust_data
cust_data = subset(cust_data, select = -c(Z_CostContact, Z_Revenue))
```

## Get rid of invalid observations I

```{r invalid_instances}
table(cust_data$Marital_Status)
knitr::kable(cust_data %>% filter(Marital_Status == "YOLO"), "html", caption = "YOLO instances") %>%  kable_styling("striped") %>% scroll_box(width = "100%")
cust_data = cust_data %>% filter(Marital_Status != "YOLO")
```

## Feature extraction I
1. Get the minimum number of household members from attributes `Marital_Status`, `Kidhome` and `Teenhome`. 
2. Get the total number of accepted offers.

```{r feature_ext}
# Calculate total number of accepted campaign, minimum number of household members.
min_num_member = function(x) {
  if (x[1]=="Married" || x[1]=="Together"){
    np = 2
  } else {
    np = 1
  }
  return(as.numeric(x[2]) + as.numeric(x[3]) + np)
}
min_num_household = apply(cust_data %>% 
                            select(Marital_Status, Kidhome, Teenhome), 
          1, min_num_member)
cust_data$min_num_household = min_num_household
cust_data = cust_data %>% mutate(tot_AcceptedCmp = AcceptedCmp1 + AcceptedCmp2 + AcceptedCmp3 + AcceptedCmp4 + AcceptedCmp5 + Response)
```


## Restructuring of data types 

```{r data_types_transform}
# Change the date from string to date format
cust_data$Dt_Customer = as.Date(cust_data$Dt_Customer, 
                                     format = c("%d-%m-%Y"))
# cust_data %>% as_tibble() %>% glimpse()
# Change the categorical features from character to factor
categorical_features = c("Education", "Marital_Status", "AcceptedCmp1",
                         "AcceptedCmp2", "AcceptedCmp3", "AcceptedCmp4",
                         "AcceptedCmp5", "Complain", "Response")
for (i in 1:length(categorical_features)){
  cust_data[[categorical_features[i]]] = 
    as.factor(cust_data[[categorical_features[i]]])
}
# ordered categorical variables: Kidhome and Teenhome
cust_data$Kidhome = factor(cust_data$Kidhome, 
                           levels = seq(from = min(cust_data$Kidhome), 
                                        to = max(cust_data$Kidhome), 
                                        by = 1), ordered = TRUE)
cust_data$Teenhome = factor(cust_data$Teenhome, 
                           levels = seq(from = min(cust_data$Teenhome), 
                                        to = max(cust_data$Teenhome), 
                                        by = 1), ordered = TRUE)
```


## Feature extraction II

Change the attribute `Year_Birth` to age and change the `Dt_Customer` to number of days a customer is enrolled with the company.
```{r feature_ext_2}
# Change Year_Birth to age
current_year = 2014
cust_data = cust_data %>% mutate(age = current_year - Year_Birth) %>% 
  subset(select = -c(Year_Birth))
# Change Dt_customer to days of enrollment. Assume that the current date is 1st July 2014
current_date = as.Date("2014-07-01")
cust_data = cust_data %>% 
  mutate(days_enroll = difftime(current_date, Dt_Customer, units = c("days")) %>% 
           as.numeric()) %>% 
  subset(select = -c(Dt_Customer))
```

## Get rid of outliers

```{r outlier_removal}
summary(cust_data$age)
cust_data = cust_data %>% filter(age < 100)
cust_data %>% arrange(desc(Income)) %>% head(3) %>% kable("html", caption = "Income in descending order") %>% kable_styling("striped") %>% scroll_box(width = "100%")
cust_data = cust_data %>% filter(ID!=9432)
```

## Deal with missing values
There are 3 types of missing values:   
1. missing completely at random (**MCAR**)
2. missing at random (**MAR**)
3. missing not at random (**MNAR**).
Refer to [this awesome article](https://www.ncbi.nlm.nih.gov/books/NBK493614/) for more info on each types of missing values. 
There are several strategies to choose from when we encounter missing values in the data:  
1. Complete case analysis (ignore observation(s) with missing values).  
2. Single imputation - mean, median, regression model.  
3. Multiple imputation - leverage distribution of observed data to estimate the missing values.  
Multiple imputation is generally superior to single imputation because the missing values are calculated multiple times with many different plausible values being pooled together to get the estimates. For this experiment, Multiple Imputation by Chained Equations (MICE) is implemented to obtain the estimation of missing `Income` instances.
```{r miss_value_func, warning=FALSE}
# Write a utility function to check for missing values.
check_missing_values = function(cust_data){
  vec_missing = apply(cust_data, 2, function(x){ sum(any(is.na(x)))})
  idx_missing = which(as.vector(vec_missing)==1)
  if (length(idx_missing)>0){
    cat("The columns(features) with missing values: ", 
      colnames(cust_data)[idx_missing])
    
  } else {
    print("No missing values")
  }
  
}
check_missing_values(cust_data)
vis_miss(cust_data)
cat(" \nThe number of missing values for income: ", sum(is.na(cust_data$Income)))
idx_missing = which(is.na(cust_data$Income))
```

### Perform multiple imputation.

```{r MICE, warning=FALSE}
# MICE implementation
imp = mice(cust_data, maxit = 0, print = F)
pred = imp$predictorMatrix
pred[,"ID"] = 0
# Use the default value of 5 iterations
imp = mice(cust_data, seed = 10, predictorMatrix = pred, print = F)
stripplot(imp, Income~.imp, pch = 20, cex = 2)
cust_data = complete(imp, 5)
# Save the preprocessed data for further analysis
saveRDS(cust_data, file = "preprocessed_cust_data.rds")
```

--- 
This wraps up the *preprocessing of the data*. Lets take a look at the results of missing `Income` imputation
```{r show_results, echo=FALSE}
knitr::kable(cust_data[idx_missing[1:5],] %>% select(ID:Income), caption = "Missing values imputed customer data")
```

# Data visualization
Data visualization is graphical representation of information and data.

## Histogram and density plot
Lets visualize the household income. There is/are customer(s) with unusually high yearly household income based on empirical observation on the scale of x axis of the histogram plot.
```{r histogram}
#hist_data = cust_data %>% ggplot(aes(x = Income)) + 
#  geom_histogram(binwidth = 10000, boundary = 0)
cust_data %>% ggplot(aes(x = Income)) + 
  geom_histogram(aes(y = stat(count)/sum(count)), binwidth = 10000, 
                 boundary = 0) +
  ylab("Relative frequency") 
# density plot
cust_data %>% ggplot(aes(x = Income)) + geom_density(fill = "lightblue", alpha = 0.5) +
  geom_vline(aes(xintercept=mean(Income)), color = "blue", linetype = 2) + geom_vline(aes(xintercept=median(Income)), color = "red", linetype = 3) + xlim(0, 2e+05) + theme_classic() + ggtitle("density plot for customer annual household income")
cust_data %>% ggplot(aes(x = Income)) + 
  geom_histogram(aes(y = stat(count)/sum(count)), binwidth = 0.2, 
                 boundary = 0) +
  scale_x_log10() +
  ylab("Relative frequency") +
  ggtitle("Distribution of income in log10 scales")
```

Most of the customers (`r 100*round(nrow(cust_data %>% filter(Income <= 1e+05))/nrow(cust_data), 3)` %) have income of equal or less than 100k. In fact, (`r 100*round((cust_data %>% filter(Income >= 2e+04, Income <= 9e+04) %>% nrow())/nrow(cust_data), 4)` %) of the customers have annual household income within 20k-90k. 

## Scatter plots among continuous predictors

```{r scatterplot_matrix_1}
# Linear correlation and scatter plots matrix for amount spent on different merchandise
idx_cont_features = grepl("Mnt", colnames(cust_data))
cust_data[, idx_cont_features] %>% ggpairs(title = "correlogram")
```

```{r scatterplot_matrix_2}
# Linear correlation and scatter plots matrix for number of purchases.
idx_cont_features = grepl("Num", colnames(cust_data))
cust_data[, idx_cont_features] %>% ggpairs(title = "correlogram")
```

## Correlation matrix / Correlogram
Linear correlation between numerical features.
```{r correlation_matrix}
seq = c("Mnt", "Num","Age", "days", "Income", "Recency", "household","tot")
a = sapply(seq, function(x) grep(x, colnames(cust_data)))
idx = c()
for (i in (1:length(seq))){
  idx = c(idx, a[[i]])
}
corr_mat = cor(cust_data[,idx])
corr_mat[upper.tri(corr_mat)] = NA
melted_cormat = melt(corr_mat, na.rm = TRUE)
# The dimension becomes nx3 matrix
melted_cormat %>% ggplot(aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") + 
  scale_fill_gradientn(limit = c(-1,1), colors = hcl.colors(40, "Earth")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 10, hjust = 1)) + coord_fixed() +
  geom_text(aes(Var1, Var2, label = round(value, 2)), color = "black", size = 1.5)
```

## Mosaic Plot
Response rate from different groups of population.
```{r mosaic_plot, warning=FALSE}
cust_data %>% ggplot() +
  geom_mosaic(aes(x = product(Marital_Status, Education), fill = Response)) +
  theme(axis.text.x = element_text(face = "bold", angle = 90),
        axis.text.y = element_text(face = "bold")) +
  ggtitle("Mosaic Plot of Response, Marital status and Education")
cust_data %>% ggplot() +
  geom_mosaic(aes(x = product(Kidhome, Teenhome), fill = Response)) +
  ggtitle("Mosaic Plot of Response, Kidhome and Teenhome")
```

## Parallel set plots
```{r parallel_set_plots}
data_group = cust_data %>% group_by(Education, Marital_Status, Kidhome, 
                                    Teenhome, Response) %>% 
  summarise(summ = n())
data_parallel = reshape2::melt(data_group)
data_parallel = gather_set_data(data_parallel, 1:5)
ggplot(data_parallel, aes(x, id = id, split = y, value = value)) +
  geom_parallel_sets(aes(fill = Response), alpha = 0.3, axis.width = 0.1) +
  geom_parallel_sets_axes(axis.width = 0.1) +
  geom_parallel_sets_labels(colour = 'white')
```
There is virtually no chance for the customers with 2 children or 2 teenagers in their household to accept the offer.


## Grouped box plots / violin plots 
Based on minimum number of household members, marital status and education. 

### Visualize income
Distribution of income for different categories of customers.
```{r violin_plots}
seq = c("Income", "Education", "Marital", "min")
idx_plot = sapply(seq, function(x) grep(x, colnames(cust_data)))
idx = c()
for (i in (1:length(seq))){
  idx = c(idx, idx_plot[[i]])
}
cust_data %>% dplyr::select(idx) %>% 
  ggplot(aes(x = Education, y = Income, fill = Education)) +
  geom_violin(scale = "count", draw_quantiles = c(0.25, 0.5, 0.75)) + stat_summary(fun = mean, geom = "point", shape=21, size=1.5, color = "red") + ylim(0, 2e+05)
cust_data %>% dplyr::select(idx) %>% dplyr::filter(!(Marital_Status %in% c("Alone", "Absurd"))) %>% 
  ggplot(aes(x = Marital_Status, y = Income, fill = Marital_Status)) +
  geom_violin(scale = "count", draw_quantiles = c(0.25, 0.5, 0.75)) + stat_summary(fun = mean, geom = "point", shape=21, size=1.5, color = "red") + ylim(0, 2e+05) + ggtitle("Distribution of household income based on marital status")
cust_data %>% dplyr::select(idx) %>% mutate(num_household = as.factor(min_num_household)) %>%
  ggplot(aes(x = num_household, y = Income, fill = num_household)) +
  geom_violin(scale = "count", draw_quantiles = c(0.25, 0.5, 0.75)) + stat_summary(fun = mean, geom = "point", shape=21, size=1.5, color = "red") + ylim(0, 2e+05) + labs(title = "Distribution of household income based on minimum number of household members", x = "minimum number of household members")
```

### Visualize expenditure
Show expenditure of different types of products.
```{r grouped_box_plots}
seq = c("Mnt", "Education", "Marital", "min")
idx_plot = sapply(seq, function(x) grep(x, colnames(cust_data)))
idx = c()
for (i in (1:length(seq))){
  idx = c(idx, idx_plot[[i]])
}
cust_data %>% dplyr::select(idx) %>% 
  pivot_longer(starts_with("Mnt"), names_to = "purchase_types",
               values_to = "amount_spent") %>% 
  ggplot(aes(x = Education, y = amount_spent)) +
  geom_boxplot(fill = "chocolate2", notch = TRUE) + stat_summary(fun = mean, geom = "point", shape=21, size=1.5, color = "red") + theme_classic() + facet_wrap(~purchase_types, ncol = 2, scales = "free_y")
cust_data %>% dplyr::filter(!(Marital_Status %in% c("Alone", "Absurd"))) %>% dplyr::select(idx) %>% 
  pivot_longer(starts_with("Mnt"), names_to = "purchase_types",
               values_to = "amount_spent") %>% 
  ggplot(aes(x = Marital_Status, y = amount_spent)) +
  geom_boxplot(fill = "chocolate2", notch = TRUE) + stat_summary(fun = mean, geom = "point", shape=21, size=1.5, color = "red") + theme_classic() + facet_wrap(~purchase_types, ncol = 2, scales = "free_y") 
cust_data %>% dplyr::select(idx) %>% mutate(num_household = as.factor(min_num_household)) %>% 
  pivot_longer(starts_with("Mnt"), names_to = "purchase_types",
               values_to = "amount_spent") %>% 
  ggplot(aes(x = num_household, y = amount_spent)) +
  geom_boxplot(fill = "chocolate2", notch = TRUE) + stat_summary(fun = mean, geom = "point", shape=21, size=1.5, color = "red") + theme_classic() + facet_wrap(~purchase_types, ncol = 2, scales = "free_y") 
```

```{r, echo=FALSE}
R.version
```
